---
title: "AI and LLMs"
subtitle: "What an LLM Actually Does"
author: "David Riegert"
engine: knitr
execute:
  echo: true
format:
  revealjs:
    theme: default
    margin: 0.05
    incremental: false
    logo: TUPMS.png
    css: style.css
    include-after-body: replay-svg.html
    pdf-separate-fragments: false
slide-number: true
smaller: false
---

```{r}
#| label: packages
#| include: false

library(tidyverse)
library(tidyr)
library(knitr)
library(kableExtra)

options(knitr.table.format = "html")  # honour escape = FALSE so <strong>/<em> in cells render
theme_set(theme_bw())
```

# What an LLM Actually Does

## The Whole Trick: Predict the Next Token {style="font-size:95%;"}

> An LLM is a function that, given the text so far, outputs a **probability distribution over what comes next** --- then samples from it. Repeat.

. . .

That's it. Everything else --- "reasoning," "writing code," "knowing facts" --- is this loop running over and over.

. . .

This is a *conditional distribution*.

$$
P(\text{next token} \mid \text{all previous tokens})
$$

## How It "Reads": Tokens {style="font-size:90%;"}

The model doesn't see words or letters --- it sees **tokens**: subword chunks.

- `"statistician"` might split into `stat` + `ist` + `ician`.
- `"unbelievable"` might be `un` + `believ` + `able`.
- Common words = one token; rare words = several.
- Numbers, whitespace, and punctuation are tokens too --- which is why models can be oddly bad at arithmetic.

## {.svgslide}

<object class="figsvg" tabindex="-1" type="image/svg+xml" data="tokens.svg" data-dur="11" data-loops="2" data-hold="0.9"></object>

## Tokenizing Examples

<div style="height:200px"></div>

**How an interaction with LLMs are tokenized** [tiktokenizer.vercel.app](https://tiktokenizer.vercel.app) 

## Tokens $\to$ a Distribution You Can Sample {style="font-size:80%;"}

The model gives every candidate next token a score which is turned into probabilities; **temperature** controls how "peaked" the distribution is:

- **Low temperature** $\to$ spiky $\to$ picks the most likely token *("greedy")*.
    - Deterministic when `temperature = 0`
- **High temperature** $\to$ flat $\to$ takes more risks *("creative" / hallucination-prone)*.

*Where do the scores come from? From data.*  
(A _lot_ of data.)

## {.svgslide}

<object class="figsvg" tabindex="-1" type="image/svg+xml" data="next_token_sampling.svg" data-dur="14" data-loops="2" data-hold="0.9"></object>

## {.svgslide}

<object class="figsvg" tabindex="-1" type="image/svg+xml" data="temperature.svg" data-dur="12" data-loops="2" data-hold="0.9"></object>

## Two Phases of Training {style="font-size:88%;"}

```{r}
#| echo: false

tibble(
  Phase = c("Pre-training", "Post-training"),
  `What happens` = c(
    "Predict the next token over ~the whole internet. Unsupervised, enormously expensive.",
    "Instruction tuning + <strong>RLHF</strong> (human/AI feedback). Cheaper, shapes behaviour."),
  Analogy = c("Learns the <em>shape</em> of language",
              "Learns to be <em>helpful and follow instructions</em>")
) |>
  kable(escape = FALSE) |>
  kable_styling(font_size = 26, full_width = FALSE) |>
  column_spec(1, bold = TRUE)
```

**Key point:** pre-training builds a raw next-token engine; post-training makes it answer *you* instead of just autocompleting.

::: {.aside}
**Fine-tuning** is post-training you do yourself: take a released model and keep training it on *your* data (your grading style, your domain) to specialize its behaviour --- far cheaper than pre-training from scratch.
:::

## "Modes": Fast vs. Reasoning {style="font-size:88%;"}

Often the same model, but a different amount of *computation per answer*:

- **Fast mode** --- answers immediately. Great for quick, low-stakes tasks.
- **Reasoning / extended mode** --- the model first generates a long internal "thinking" trace (more tokens, more time, more cost) before answering. Better on multi-step problems.

. . .

We often do not see the generated intermediate tokens (the "thinking trace"). What we see is the final output.

## Mechanical Takeaway {style="font-size:92%;"}


LLMs **generates**.<br>It does not **retrieve**.

. . .

A fluent answer is a *high-probability* answer --- **not** a *true* one. That gap is why we have to be careful when evaluating what an LLM gives us for a response.

. . .

```{r}
#| echo: false

tibble(
  ` ` = c("Fluent &amp; <strong>wrong</strong>", "Fluent &amp; <strong>right</strong>"),
  `What it looks like` = c(
    "\"A 95% confidence interval means there's a 95% chance the true parameter lies within the interval.\"",
    "\"A 95% confidence interval is an interval procedure that, across repeated sampling, captures the true parameter 95% of the time.\""),
  `What it is` = c(
    "Statistically <em>incorrect</em>, but reads perfectly",
    "Statistically <em>correct</em>, reads perfectly")
) |>
  kable(escape = FALSE) |>
  kable_styling(font_size = 22, full_width = FALSE) |>
  column_spec(1, bold = TRUE)
```
