---
title: "AI and LLMs"
subtitle: "AI Fluency: The 4D Framework"
author: "David Riegert"
engine: knitr
execute:
  echo: true
format:
  revealjs:
    theme: default
    margin: 0.05
    incremental: false
    logo: TUPMS.png
    css: style.css
    include-after-body: replay-svg.html
    pdf-separate-fragments: false
slide-number: true
smaller: false
---

```{r}
#| label: packages
#| include: false

library(tidyverse)
library(tidyr)
library(knitr)
library(kableExtra)

options(knitr.table.format = "html")  # honour escape = FALSE so <strong>/<em> in cells render
theme_set(theme_bw())
```

# AI Fluency: The 4D Framework

## What Is AI Fluency? {style="font-size:95%;"}

> **AI Fluency** is "the ability to effectively, ethically, and safely understand, evaluate, and collaborate with artificial intelligence systems."

. . .

It is **not** about getting better answers faster --- it is about **building better judgment**.

. . .

AI Fluency means interacting with AI systems in ways that are **effective, efficient, ethical, and safe** --- regardless of which models or tools emerge.

::: {.aside}
Framework: Rick Dakan & Joseph Feller, with Anthropic (2025). Released under CC BY-NC-SA 4.0. Companion courses: *AI Fluency: Framework & Foundations* and *AI Fluency for Students* (Anthropic Academy, free).
:::

## The 4D Framework {style="font-size:90%;"}

```{r}
#| echo: false

tibble(
  D = c("Delegation", "Description", "Discernment", "Diligence"),
  Question = c(
    "Should AI do this --- and how much of it?",
    "Do I know how to ask for what I want?",
    "How good is what I got back, really?",
    "Who is accountable for the outcome?"),
  Official = c(
    "Setting goals and deciding whether, when and how to engage with AI.",
    "Effectively describing goals to prompt useful AI behaviors and outputs.",
    "Accurately assessing the usefulness of AI outputs and behaviors.",
    "Taking responsibility for what we do with AI and how we do it.")
) |>
  kable(escape = FALSE) |>
  kable_styling(font_size = 22, full_width = FALSE) |>
  column_spec(1, bold = TRUE)
```

Four **interconnected** competencies --- each is a collection of skills, knowledge, insights, and values, not a checkbox.

## The Four Ds in Practice {style="font-size:88%;"}

```{r}
#| echo: false

tibble(
  D = c("Delegation", "Description", "Discernment", "Diligence"),
  `In our context` = c(
    "Grading: AI drafts feedback, human owns the judgment.",
    "A prompt is a conditional distribution --- audience, length, format, constraints.",
    "Fluent $\neq$ true: audit before you trust (that whole morning session).",
    "Your name is on the work --- disclose and stand behind AI-assisted output.")
) |>
  kable(escape = FALSE) |>
  kable_styling(font_size = 24, full_width = FALSE) |>
  column_spec(1, bold = TRUE)
```

# Delegation

## D1 --- Delegation {style="font-size:90%;"}

**Setting goals and deciding whether, when, and how to engage with AI.**

- Understand your **goal** before touching the tool.
- Know the **capability envelope**: what AI is good, weak, or forbidden at.
- Make a *strategic choice* about the split between human and machine work --- **not** "do it all myself" or "do it all for me".

::: {.fragment}
**Default-to-automation is the failure mode:** asking AI to do the whole task because it's easiest. That's the concern behind every academic-integrity policy --- and the reason delegation is D1.
:::

# Description

## D2 --- Description {style="font-size:90%;"}

**Effectively describing goals to prompt useful AI behaviors and outputs.**

Three sub-skills:

- **Product description** --- what you want, in terms of *output, format, audience, style*.
- **Process description** --- how you want the work done (steps, constraints, "don'ts").
- **Behavior description** --- how you want the AI to *act* (persona, rules, escalation).

. . .

This is the **Prompting** session made formal: every lever from that talk (clear instructions, few-shot examples, system roles) is Description competence.

# Discernment

## D3 --- Discernment {style="font-size:90%;"}

**Accurately assessing the usefulness of AI outputs and behaviors.**

Remember part 1: a fluent answer is a *high-probability* answer --- **not** a *true* one.

. . .

Discernment in practice:

- **Verify, don't just read** --- especially numbers, citations, and statistical claims.
- **Artifact effect**: the more polished the output, the *less* people check it --- fluency is a sedative.
- **Iteration**: refine, don't settle --- the single strongest correlate of good AI use in Anthropic's data.

::: {.aside}
This is where the **audit-literacy** session (session 2) plugs in: checking output is the practiced form of Discernment.
:::

# Diligence

## D4 --- Diligence {style="font-size:90%;"}

**Taking responsibility for what we do with AI and how we do it.**

- **Accountability** --- you own the output. "The model did it" is not a defense.
- **Transparency** --- disclose AI involvement where it matters (assignments, publications, student work).
- **Ethics** --- data privacy, IP, fairness: *how* you do it with AI counts as much as *what* you do.

. . .

The other three Ds make AI **useful**; Diligence keeps it **legitimate**.

# Three Modes of Interaction

## Automation, Augmentation, Agency {style="font-size:88%;"}

The framework's second axis: **how** you hand work over.

```{r}
#| echo: false

tibble(
  Mode = c("Automation", "Augmentation", "Agency"),
  Definition = c(
    "AI **executes** specific tasks based on human instruction.",
    "Humans and AI **collaborate** as thinking partners.",
    "Humans **configure** AI to independently perform tasks on their behalf.")
) |>
  kable(escape = FALSE) |>
  kable_styling(font_size = 26, full_width = FALSE) |>
  column_spec(1, bold = TRUE)
```

::: {.aside}
Notice the gradient: more capability, **more autonomy** --- and, per the audit ladder, more checking required at every step up.
:::

# Where This Fits

## 4D $\times$ the Rest of the Day {style="font-size:90%;"}

The 4D framework organizes **how well** you work with AI. The 3D Autonomy Ladder from session 1 organizes **how much** you hand off.

- **Delegation** $\leftrightarrow$ choosing your rung on the autonomy ladder
- **Description** $\leftrightarrow$ the prompting session (how to condition the distribution)
- **Discernment** $\leftrightarrow$ the audit-literacy session (checking, not trusting)
- **Diligence** $\leftrightarrow$ the guardrails: policy, disclosure, your own standards

. . .

**Takeaway:** the tools and models will keep changing. Fluency is the part that doesn't --- and it is what you leave this day with.
